\chapter{Background}

\section{Autonomous Drone Systems}

Autonomous drone systems are becoming increasingly prominent in applications that are otherwise time-consuming or dangerous using traditional methods.
Such applications include infrastructure inspection, for example monitoring bridge structures \cite{ellenberg2016bridge}, environmental monitoring such
as wildlife surveys \cite{anderson2013uav}, and search and rescue (SAR) missions, where drones can assist in locating people after disasters \cite{erdelj2016uav}.
Unmanned Aerial Vehicles (UAVs) are also widely used in surveillance and security applications, including border monitoring and crowd observation \cite{zhou2023uav}.
In addition, drones have been successfully deployed for logistics and delivery tasks, such as Amazon Prime Air \cite{amazon_prime_air}, as well as real-world medical
 transport solutions, including long-range delivery of blood samples in Denmark \cite{investindk2022health_drone}.
Large-scale industrial and energy infrastructures can likewise be efficiently monitored using UAVs \cite{nooralishahi2021drone}.

These diverse applications highlight the potential of UAVs; however, understanding the characteristics and capabilities of different drone platforms is essential.

Multirotor drones are widely used due to their ability to hover, perform vertical takeoff and landing (VTOL), and navigate precisely in constrained environments.
These capabilities make them particularly suitable for urban operations such as inspection and delivery.
In contrast, fixed-wing drones are designed for long-endurance missions and can cover large areas efficiently, making them well suited for large-scale mapping and
environmental monitoring tasks. However, they lack hovering capability and require more space for takeoff and landing, limiting their use in high-precision or urban scenarios.

The suitability of a drone for a given mission depends heavily on the payload it carries.
RGB cameras provide visual data for inspection, surveillance, and mapping tasks, and can also be used for motion estimation through visual odometry \cite{rostum2023visual}.
Thermal cameras mounted on UAVs have been successfully applied in infrastructure inspection, for example to detect subsurface delamination in bridge decks using infrared 
thermography \cite{ellenberg2016bridge}.
LiDAR-equipped UAVs enable accurate three-dimensional mapping of environments, for instance in forestry applications where canopy structure can be estimated from aerial point 
clouds \cite{wallace2016assessment}.
In addition, onboard sensors such as Inertial Measurement Units (IMUs), GPS, and barometers are used for state estimation, including position, velocity, and altitude.

In multi-drone systems, payload diversity introduces functional heterogeneity, meaning that not all drones are capable of performing all tasks. 
This has direct implications for task allocation, as tasks must be matched with drones that possess the required sensing and operational capabilities.

Many modern UAV applications rely on artificial intelligence (AI) for tasks such as perception, navigation, and decision-making. 
However, developers often face strict hardware constraints in drone systems. While cloud-based AI can provide significant computational resources, it introduces latency and 
dependence on network availability.
The emergence of edge AI enables real-time decision-making by allowing drones to process sensor data locally and operate independently of external infrastructure
\cite{mcenroe2022edgeaiuav}. These platforms often support hardware acceleration through GPUs, DSPs, or NPUs, enabling efficient execution of machine learning models.
This is particularly relevant for deploying compact and quantized large language models (LLMs) directly onboard drones.

\section{Multi-Agent Systems}

Multi-agent systems (MAS) consist of multiple autonomous agents that interact in a shared environment to achieve a collective goal. 
In the context of autonomous drone systems each drone can be modelled as an autonomous agent with it's own sensing and decision making capabilities.
Deploying heterogeneous UAV agents increases the number of available data modalities, the area that the drone swarm can inspect and provides backup options in case of
drone failures, thereby increasing mission success rate in constrast to a single agent solution \cite{pantelimon2019survey}.
The key challenges in multi-agent systems are the efficient coordination between agents and the creation of a mission plan for heterogeneous agents,
particularly under constraints such as limited communication, energy, and computation.

\section{Multi-Agent Systems}

Multi-agent systems (MAS) consist of multiple autonomous agents that interact within a shared environment to achieve individual or collective goals. 
In the context of autonomous drone systems, each drone can be modeled as an autonomous agent with its own sensing, computation, communication, and decision-making capabilities. 
Deploying heterogeneous UAV agents can increase the number of available sensing modalities, expand the area that can be inspected, and provide redundancy in case of drone failures, 
thereby improving mission robustness compared to a single-agent solution \cite{pantelimon2019survey}. 
However, multi-agent systems also introduce challenges related to coordination, task allocation, and mission planning, 
particularly under constraints such as limited communication bandwidth, finite battery capacity, and restricted onboard computation.

\subsection{Coordination in Multi-Agent Systems}

Two common architectural approaches for coordination in multi-agent systems are centralized and decentralized coordination. 
In a centralized architecture, a base station or ground control station is responsible for critical mission decisions and coordinates the behavior of the agents. 
Since the central unit is often equipped with greater computational resources than the individual agents, centralized approaches can support more complex planning and optimization. 
However, this architecture introduces a single point of failure: if the base station fails or communication with it is lost, the mission may be compromised.

In a decentralized architecture, each agent is capable of local decision-making and may communicate directly with other agents. 
This can improve robustness, since the system can remain operational even if individual agents fail. 
Decentralized systems may also scale better as new agents are added, since control is not concentrated in a single central unit. 
However, coordination becomes more challenging because agents must reach decisions using local information and distributed communication mechanisms.

Several mechanisms can support decentralized coordination. 
Leader election methods dynamically select a coordinating agent responsible for organizing group decisions. 
Auction-based methods allocate tasks by allowing agents to bid based on local cost, capability, or availability, with tasks assigned according to the most suitable bid. 
Voting mechanisms allow agents to collectively choose between alternatives, 
while bargaining and negotiation approaches enable agents to iteratively exchange proposals until an agreement is reached.

Hybrid approaches combine elements of both centralized and decentralized coordination. 
For example, a ground station may perform high-level mission planning, 
while individual UAVs retain autonomy to make local decisions based on their current state and surrounding environment. 
Such architectures can exploit the computational resources of centralized systems while preserving some of the robustness and adaptability of decentralized systems.

\subsection{Planning in Multi-Agent Systems}

\section{Large Language Models}

\subsection{Transformer Architecture}
attention mechanism, quantized LLMs
\subsection{Prompting and Instruction Following}
prompting techniques, instruction following models, roles (human/assistant)
\subsection{LLMs for Planning and Reasoning}
chain of thought, structured outputs (JSON)
limitations: hallucinations, inconsistency

\section{System Constraints in Autonomous Drone Systems}
Latency and Computation Constraints

Edge vs Cloud Processing

Robustness and Failure Handling


\section{Related Work}

\subsection{LLMs in Robotics and Autonomous Systems}
Include research gap in each subsection.
\subsection{Multi-Agent Task Allocation and Coordination}
\subsection{Replanning and Adaptive Systems}